How does an Experimental Software Engineering Lab Work? An Ethnographic Study

### 2023-07-25 Article.tex
Corregido el problema de la bibliografia y de las listas itemize
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

%%%%%%Begin MyPackages%%
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{framed}
\usepackage{multirow}
\usepackage{color}
\usepackage{tabulary}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{url}
\usepackage{soul}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{caption}
\usepackage[title]{appendix}
\usepackage{ulem}
\usepackage{placeins}
\usepackage{balance}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
%%%%%%end MyPackages%%%%

@@

%%%%%%% Begin Content %%%%%%%
\input{tex/introduction}
\input{tex/research-method}
\input{tex/research-execution}
\input{tex/results}
\input{tex/discussion}
\input{tex/threats-to-validity}
\input{tex/conclusion}
\input{tex/data-availability-statements}
%%%%%%% End Content %%%%%%%%%

@@

\section*{Acknowledgment}

Insert the Acknowledgment text here.

@@

\bibliographystyle{imaiai}
\bibliography{bibliography}
%
% once the .bbl file has been generated then place the text in your article.

>>>>

%%%%%%Begin MyPackages%%
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{framed}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{color}
\usepackage{tabulary}
\usepackage{enumitem}
	\setlist[itemize]{leftmargin=5.5mm}

%\usepackage{hyperref}
%\usepackage{url}
\usepackage[hyphens]{url}
\usepackage[hidelinks]{hyperref}
\hypersetup{breaklinks=true}
\urlstyle{same}
%\usepackage{cite}
\usepackage{soul}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{caption}
\usepackage[title]{appendix}
\usepackage{ulem}
\usepackage{placeins}
\usepackage{balance}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
%%%%%%end MyPackages%%%%

@@

%%%%%%% Begin Content %%%%%%%
\input{tex/introduction}
\input{tex/research-method}
\input{tex/research-execution}
\input{tex/results}
\input{tex/discussion}
\input{tex/threats-to-validity}
\input{tex/conclusion}
%\input{tex/data-availability-statements}
%%%%%%% End Content %%%%%%%%%

@@

\section*{Acknowledgment}
This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).

@@

\Urlmuskip=0mu plus 1mu\relax
\bibliographystyle{iet}
\bibliography{bibliography}
%
% once the .bbl file has been generated then place the text in your article.

### 2023-07-25 tex/abstract.tex
Last changes
These are the last changes previous the submission

\begin{abstract}
\textbf{Context:} Replication is complex in Experimental Software Engineering (ESE). Replication packages and families of experiments have been applied with limited success. We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, or intrinsic reasons, i.e., the existing replication procedures do not fulfill the researchers' needs.
\textbf{Objective:} Find out how ESE researchers conduct experiments in practice.
\textbf{Method:} Ethnographical study with an experimental research group.
\textbf{Results:} We have created conceptual and process models representing experimentation, replication, and synthesis in the target research group. These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels. The experimental process carried out in the group differs from folk knowledge and textbooks in (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts, and (4) the viewpoints that different sub-areas or families of experiments have about the experimental process.
\textbf{Conclusions:} The differences between the actual lab processes vs. the idealized versions probably harm knowledge transfer and make replication harder. We will conduct further research with more research groups, e.g., using a survey to explore whether our findings have general applicability.
\end{abstract}

>>>>

\begin{abstract}
\textbf{Context:} Replication is complex in Experimental Software Engineering (ESE). Replication packages and families of experiments have been applied with limited success. We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, or intrinsic reasons, i.e., the existing replication procedures do not fulfill the researchers' needs.
\textbf{Objective:} Determine how ESE researchers conduct experiments in practice.
\textbf{Method:} Ethnographical study with an experimental research group.
\textbf{Results:} We have created conceptual and process models representing experimentation, replication, and synthesis in the target research group. These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels. The experimental process carried out in the group differs from folk knowledge and textbooks in (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts, and (4) the viewpoints that different subareas or families of experiments have about the experimental process.
\textbf{Conclusions:} The differences between the actual lab processes vs. the idealized versions probably harm knowledge transfer and make replication harder. We will conduct further research with more research groups, e.g., using a survey to explore whether our findings have general applicability.
\end{abstract}

### 2023-07-25 tex/conclusion.tex
Last changes
These are the last changes previous the submission

\section{Conclusions}\label{sec-conclusions}
Experimental researchers know that experimentation, as described in textbooks and articles, is only an approximation to the daily work in the labs and the field. This study shows the deviations found in a concrete experimental research group. Such divergences do not seem arbitrary but necessary from a systemic viewpoint, e.g., each people have distinct abilities and perform different tasks in a research project.

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group in which we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the labs and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k} \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests at a relatively superficial level, e.g., when researchers from different experimental families discuss their mental models.

@@

We believe that the "lab mythology" is worth studying. If the differential characteristics of the theoretical vs. actual experiment process are research-group specific, the inquiry will end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to the improvement of experimental practice.

We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring some issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic, because they require too much time) on this topic soon.

>>>>

\section{Conclusions}\label{sec-conclusions}
Experimental researchers know that experimentation, as described in textbooks and articles, only approximates the daily work in the lab and the field. This study shows the deviations found in a concrete experimental research group. Such divergences do not seem arbitrary but are necessary from a systemic viewpoint, e.g., people have distinct abilities and perform different tasks in a research project.

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group where we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the lab and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests relatively superficially, e.g., when researchers from different experimental families discuss their mental models.

@@

We believe that the "lab mythology" is worth studying. If the differential characteristics of the theoretical vs. actual experiment process are specific to the research group, the inquiry will end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to improving experimental practice.

We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

### 2023-07-25 tex/discussion.tex
Last changes
These are the last changes previous the submission

\section{Discussion and related work}\label{sec-related}
This research aims to find out how experimenters work in practice. After conducting the ethnographic research, we obtained models (process, workflow, conceptual). The conceptual model connects with empirical researchers' "prominent picture" of their research area. The process and workflow models represent a novel contribution; to our knowledge, such detailed descriptions of the experimental process are not available in the existing literature.

The models have been described in Section~\ref{sec-results} and are available in full detail online. In this section, we would like to focus on some interesting models features:

@@

		\item management (resources, research lines, etc.),

@@

		\item publication, etc.

@@

Although all empirical researchers are aware of these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature. To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation} \textbf{do not address these activities beyond marginal notes}. Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.
	\item The ethnographic study revealed one consequence of such diversity. It is possible, albeit unlikely, that one experimenter performs all experimental activities. These activities are not only complex: they demand time. In a project (research or not), the diversity of activities is usually associated with the existence of roles with well-defined responsibilities. \textbf{Also in experimentation. We identified three roles}: (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}. These roles are probably related to the RGUS' size; smaller (or larger) groups may have more generic/specialized roles. The concept of~\textquotedblleft role\textquotedblright~applied to experimentation is unusual but not entirely new. Mohamed et al. \cite{Mohamed-1993-roles-ESE} proposed a framework to conduct ESE experiments. This framework was based on a merge between statistical and SE process concepts, \textbf{including explicit roles}.
	\item The high-level conceptual model (see this \href{https://zenodo.org/record/7102405#.YyxvFOzMLUK}{\ul{link}}) could be endorsed by any experimenter (with some complaints depending on their specialization area, as it happened during the ethnography). However, when the concept of role appears in the scene, the high-level model is just \textbf{an abstraction of the information handled by the roles}. In other words: the high-level model is incomplete from the roles' viewpoint and thus partially valid only. The high-level model evokes the viewpoint of the \textit{senior experimenter} (see this \href{https://zenodo.org/record/7102464#.YyxvsezMLUK}{\ul{link}}), but with fewer details. In turn, the \textit{experiment manager} (see this \href{https://zenodo.org/record/7102450#.Yyxv6ezMLUL}{\ul{link}}) and the \textit{research manager} (see this \href{https://zenodo.org/record/7102431#.YyxwGezMLUL}{\ul{link}}) pay little attention to the design and execution details because the \textit{senior experimenter} takes care of that. Likewise, the \textit{experiment manager} is not affected by the experimental research in its global context, e.g., collaboration with other groups and replication. Such a concern fits the \textit{research manager}'s role. 

@@

There is, apparently, an underlying dichotomy between theory and practice. In theory, formal guides, e.g., replication packages, should describe the process step by step. However, the practice shows that such mechanisms cannot cope with the existing diversity of experimental protocols. In our opinion, the replication process will likely benefit from the specialized, fine-grained knowledge \textbf{at the family level}.
\end{itemize}

The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement. Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE}, \cite{Sjoberg-2005-survey-experiments-SE}, \cite{Shull-2004-Knowledge-sharing-issues-SE}. We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character. Instead, we aimed to observe how experimenters work, and we found out that they perform tasks other than the ones portrayed in textbooks. \textbf{This is our essential contribution}.

We are also aware that models represent \textbf{the point of view of a single experimental research group}. We must extend the inquiry to other experimental research groups, ideally to large portions of the ESE community, and check whether they reveal similar behavior patterns.

>>>>

\section{Discussion and related work}\label{sec-related}
This research aims to find out how experimenters work in practice. After conducting the ethnographic research, we obtained models (process, workflow, and conceptual). The conceptual model connects with empirical researchers' "prominent picture" of their research area. The process and workflow models represent a novel contribution; to our knowledge, such detailed descriptions of the experimental process are unavailable in the existing literature.

The models described in Section~\ref{sec-results} are available in full detail online. In this section, we would like to focus on some interesting model features:

@@

			\item management (resources, research lines, among others.),

@@

			\item publication, among others.

@@

Although all empirical researchers know these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature. To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}. Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.
	\item The ethnographic study revealed one consequence of such diversity. It is possible, albeit unlikely, that one experimenter performs all experimental activities. These activities are not only complex: they demand time. In a project (research or not), the diversity of activities is usually associated with roles with well-defined responsibilities. \textbf{Also in experimentation. We identified three roles}: (1) \textit{Research Manager}, (2) \textit{Experiment Manager}, and (3) \textit{Senior Experimenter}. These roles are probably related to the RGUS' size; smaller (or larger) groups may have more generic/specialized roles. The concept of~\textquotedblleft role\textquotedblright~applied to experimentation is unusual but not entirely new. Mohamed et al. \cite{Mohamed-1993-roles-ESE} proposed a framework to conduct ESE experiments. This framework was based on a merge between statistical and SE process concepts, \textbf{including explicit roles}.
	\item The high-level conceptual model (see this \href{https://zenodo.org/record/7102405#.YyxvFOzMLUK}{\ul{link}}) could be endorsed by any experimenter (with some complaints depending on their specialization area, as it happened during the ethnography). However, when the concept of a role appears in the scene, the high-level model is just \textbf{an abstraction of the information handled by the roles}. In other words: the high-level model is incomplete from the roles' viewpoint and thus partially valid. The high-level model evokes the viewpoint of the \textit{senior experimenter} (see this \href{https://zenodo.org/record/7102464#.YyxvsezMLUK}{\ul{link}}), but with fewer details. In turn, the \textit{experiment manager} (see this \href{https://zenodo.org/record/7102450#.Yyxv6ezMLUL}{\ul{link}}), and the \textit{research manager} (see this \href{https://zenodo.org/record/7102431#.YyxwGezMLUL}{\ul{link}}) pay little attention to the design and execution details because the \textit{senior experimenter} takes care of that. Likewise, the \textit{experiment manager} is not affected by the experimental research in its global context, e.g., collaboration with other groups and replication. Such a concern fits the \textit{research manager}'s role. 

@@

There is, apparently, an underlying dichotomy between theory and practice. In theory, formal guides, e.g., replication packages, should describe the process. However, the practice shows that such mechanisms cannot cope with the existing diversity of experimental protocols. The replication process will likely benefit from the specialized, fine-grained knowledge \textbf{at the family level}.
\end{itemize}

The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement. Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE}, \cite{Sjoberg-2005-survey-experiments-SE}, \cite{Shull-2004-Knowledge-sharing-issues-SE}. We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character. Instead, we aimed to observe how experimenters work and found out that they perform tasks other than those portrayed in textbooks. \textbf{This is our essential contribution}.

We also know that models represent \textbf{the point of view of a single experimental research group}. We must extend the inquiry to other experimental research groups, ideally to large portions of the ESE community, and check whether they reveal similar behavior patterns.

### 2023-07-25 tex/introduction.tex
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

Several reasons prevent replication. They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE}. In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.

Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments. Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE} \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data-generation processes vary.

This paper aims to find out \textbf{how experimental researchers conduct experiments in practice}. We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different sub-areas or families of experiments have about the overall process.

@@

The remainder of the article is structured as follows: Section \ref{sec-research-method} describes the research method. Section \ref{sec-reseach-execution} reports the stages of the ethnographic research and the observations made. Section \ref{sec-results} reports the concept and process models. Section \ref{sec-related} discusses the results and the connections with the existing literature. Section \ref{sec-threats} details the threats to validity and the related mitigation strategies. Finally, in section \ref{sec-conclusions}, we present the conclusions and possible future work.

>>>>

Several reasons prevent replication. They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments}, \cite{Demagalhaes-2015-replications-SE}. In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.

Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments. Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE}, \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.

This paper aims to determine \textbf{how experimental researchers conduct experiments in practice}. We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.

@@

The remainder of the article is structured as follows: Section \ref{sec-research-method} describes the research method. Section \ref{sec-reseach-execution} reports the stages of the ethnographic research and the observations made. Section \ref{sec-results} reports the concept and process models. Section \ref{sec-related} discusses the results and the connections with the existing literature. Section \ref{sec-threats} details the threats to validity and the related mitigation strategies. Finally, in Section \ref{sec-conclusions}, we present the conclusions and future work.
